---
title: "Lake Water Quality Analysis"
author: "Matthew Ross, completed by Samantha Clark"
date: "2/23/2022"
output: html_document
---



```{r setup, include=FALSE}
library(tidyverse) # Tidy packages
library(sf) #Spatial package that can read and create shapefiles 
library(mapview) #Interactive maps
library(LAGOSNE) #Lots and lots of clean lake data
library(USAboundaries) #USA states and counties
library(lubridate) #For dealing with date and time
```


# LAGOS Analysis


## Loading in data


### First download and then specifically grab the locus (or site lat longs)
```{r data-read}
#Lagos download script
#lagosne_get(dest_folder = LAGOSNE:::lagos_path(),overwrite=T)

#Load in lagos
lagos <- lagosne_load()


#Grab the lake centroid info
lake_centers <- lagos$locus

# Make an sf object 
spatial_lakes <- st_as_sf(lake_centers,coords=c('nhd_long','nhd_lat'),
                          crs=4326)

#Grab the water quality data
nutr <- lagos$epi_nutr

#Look at column names
#names(nutr)

```

### Subset columns nutr to only keep key info that we want


```{r}
clarity_only <- nutr %>%
  select(lagoslakeid,sampledate,chla,doc,secchi) %>%
  mutate(sampledate = as.character(sampledate) %>% ymd(.))

```


### Keep sites with at least 200 observations 

```{r}

#Look at the number of rows of dataset
#nrow(clarity_only)

chla_secchi <- clarity_only %>%
  filter(!is.na(chla),
         !is.na(secchi))

# How many observatiosn did we lose?
# nrow(clarity_only) - nrow(chla_secchi)


# Keep only the lakes with at least 200 observations of secchi and chla
chla_secchi_200 <- chla_secchi %>%
  group_by(lagoslakeid) %>%
  mutate(count = n()) %>%
  filter(count > 200)


```


### Join water quality data to spatial data

```{r}
spatial_200 <- inner_join(spatial_lakes,chla_secchi_200 %>%
                            distinct(lagoslakeid,.keep_all=T),
                          by='lagoslakeid')


```

### Mean Chl_a map

```{r}
### Take the mean chl_a and secchi by lake

mean_values_200 <- chla_secchi_200 %>%
  # Take summary by lake id
  group_by(lagoslakeid) %>%
  # take mean chl_a per lake id
  summarize(mean_chl = mean(chla,na.rm=T),
            mean_secchi=mean(secchi,na.rm=T)) %>%
  #Get rid of NAs
  filter(!is.na(mean_chl),
         !is.na(mean_secchi)) %>%
  # Take the log base 10 of the mean_chl
  mutate(log10_mean_chl = log10(mean_chl))

#Join datasets
mean_spatial <- inner_join(spatial_lakes,mean_values_200,
                          by='lagoslakeid') 

#Make a map
mapview(mean_spatial,zcol='log10_mean_chl')
```


# Class work

## 1) What is the correlation between Secchi Disk Depth and Chlorophyll a for sites with at least 200 observations?

- Here, I just want a plot of chla vs secchi for all sites 

```{r}
#plot
ggplot(chla_secchi_200) + aes(x=chla, y=secchi) + geom_point() + geom_smooth()
```

As the amount of chla increases, secchi decreases.

## Why might this be the case? 
  As there is greater chla, there is more algae in the water. The more algae, the less clarity and therefore a lower secchi number.

## 2) What states have the most data? 

### 2a) First you will need to make a lagos spatial dataset that has the total number of counts per site.

```{r}

countoflakes <- nutr %>%
  group_by(lagoslakeid) %>%
  summarize(count = n())

twoa <- inner_join(spatial_lakes, countoflakes %>%
                            distinct(lagoslakeid,.keep_all=T),
                          by='lagoslakeid')
```


### 2b) Second, you will need to join this point dataset to the us_boundaries data. 

```{r}

states <- us_states()
lakesbystate <- st_join(twoa, states)

```


### 2c) Then you will want to group by state and sum all the observations in that state and arrange that data from most to least toatl observations per state. 

```{r}

Final <- lakesbystate %>%
  group_by(state_name) %>%
  summarize(
    sum = sum(count)
  ) %>%
  arrange(desc(sum))
```

The states with the most data are Minnesota (358137), Wisconsin (145910), and Michigan (100683).
The next largest are Maine, New York, Vermont, Rhode Island, Missouri, and New Hampshire with between 90,000 and 10,000 lakes. 

##3 Is there a spatial pattern in Secchi disk depth for lakes with at least 200 
observations?

```{r}

#filter out info we dont need
clarity_only2 <- nutr %>%
  select(lagoslakeid,sampledate,doc,secchi) %>%
  mutate(sampledate = as.character(sampledate) %>% ymd(.))

#filter NAs
secchi <- clarity_only2 %>%
  filter(!is.na(secchi))

#filter out anything without 200 obs
secchi_200 <- secchi %>%
  group_by(lagoslakeid) %>%
  mutate(count = n()) %>%
  filter(count > 200)

### Take the mean secchi by lake
mean_secchi_200 <- secchi_200 %>%
  # Take summary by lake id
  group_by(lagoslakeid) %>%
  # take mean of secchi per lake id
  summarize(mean_secchi=mean(secchi,na.rm=T)) %>%
  #Get rid of NAs
  filter(!is.na(mean_secchi))

#Join datasets
mean_spatial_secchi <- inner_join(spatial_lakes,mean_secchi_200,
                          by='lagoslakeid') 

#Make a map
mapview(mean_spatial_secchi)

```

The majority of lakes at least 200 observations are in the upper north east of the US. Specifically, in Minnesota, Michigan, and Maine.
